Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-07 HickernellMCQMC2024_QMCTutorial.tex
M2 Mac updates

%%Fred's temporary macros

@@

\keywords{}
\end{abstract}
%
\section{Introduction} \label{sec:intro}
There are may settings where key underlying quantities that affect the outcome are unknown, e.g.,
\begin{itemize}
	\item Future market forces, which affect financial risk,
	\item The porosity of rock, which affect the extraction of oil or gas, or
	\item Elemetary particle interactions in a high energy physics experiment, which affect what is observed.
\end{itemize}
In such situations the unknown inputs are often modeled using random variables or stochastic processes.  Computations are performed by generating a multidude of possible scenarios informed by the assumed probability distribution of the input. These are used to estimate the mean, quantiles, and/or probability distribution of the outcome.  This is the essence of the Monte Carlo method.

@@

How one generates scenarios
%
\section{Where in Practice} \label{sec:practice}
\section{Low Discrepancy Sequence Constructions} \label{sec:construct}
\section{Discrepancy} \label{sec:discrepancy}
\section{Stopping Criteria} \label{sec:stop}
\section{Reformulating Our Problem} \label{sec:reformulate}
\section{Ongoing Research} \label{sec:onging}
\section{Conclusion} \label{sec:conclusion}

>>>>

%%Fred's temporary macros
\newcommand{\figpath}{Figures}

@@

\keywords{}
\end{abstract}
%
\section{Introduction} \label{sec:intro}
There are may settings where key underlying quantities that affect the outcome are unknown, e.g.,
\begin{itemize}
	\item Future market forces, which affect financial risk,
	\item The porosity of rock, which affect the extraction of oil or gas, or
	\item Elemetary particle interactions in a high energy physics experiment, which affect what is observed.
\end{itemize}
In such situations the unknown inputs are often modeled using random variables or stochastic processes.  Computations are performed by generating a multidude of possible scenarios informed by the assumed probability distribution of the input. These are used to estimate the mean, quantiles, and/or probability distribution of the outcome.  This is the essence of the Monte Carlo (MC) method.

@@

Simple or independent and identically distributed (IID)  Monte Carlo chooses the sequence $\bsX_0, \bsX_1, \ldots$ to be IID.  Quasi-Monte Carlo (qMC) methods choose $\bsX_0, \bsX_1, \ldots$ \emph{not} IID, but with an empirical distribution that approximates the true probability distribution of $\bsX$.  The difference between these two distributions is called a \emph{discrepancy}, and the sequences used in qMC are called low discrepancy sequences.

This tutorial describes what qMC is, why we would want to use qMC, and how qMC can be implemented well.  The next session provides an illustration of qMC in practice.  This is followed by a description of how low discrepancy sequences are constructed and various measures of quality.  We then explain how randomization can improve qMC, how to decide what sample size, $n$, is sufficient, and how to rewrite the problem of interest in a qMC-friendly way.

%
\section{An Illustration of Quasi-Monte Carlo} \label{sec:practice}

To illustrate the benefits of qMC, consider the example from Keister \cite[(13)]{Kei96}, motivated by computational physics:

\begin{equation}\label{eq:keisterA}
	\mu := \int_{\mathbb{R}^d} \cos(\lVert \bst \rVert) \exp(-\lVert \bst \rVert^2) \, \rmd \bst,
\end{equation}
where $\lVert \bst \rVert := \sqrt{t_1^2 + \cdots + t_d^2}$.  This integral may be evaluated numerically by re-writing it in spherical coordinates as
\begin{equation}\label{eq:keisterExact}
	\mu = \frac{2 \pi^{d/2}}{\Gamma(d/2)}\int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r,
\end{equation}
where $2 \pi^{d/2}/\Gamma(d/2)$ is the surface area of the sphere in $d$ dimensions, and $\Gamma$ is the Gamma function.  Having a non-trivial test case with a value that can be easily calculated allows us to compute the numerical errors of various cubature schemes.

The aim is approximate $\mu$ by a sample mean,
\begin{equation} \label{eq:sample_mean}
	\hat{\mu}_n := \frac 1n \sum_{i=1}^n f(\boldsymbol{X}_i)
\end{equation}
but to do so requires some further preparation.  This integral above, $\mu$, may be thought of as the expectation of $Y := g(\bsT) := \pi^{d/2} \cos(\lVert \bsT \rVert)$, where $T_1, \ldots, T_d$ are IID Gaussian random variables with zero mean and variance $1/2$, i.e., $\bsT \sim \calN(\bszero,\mathsf{I}/2)$:
\begin{equation}\label{eq:keisterB}
	\mu = \int_{\mathbb{R}^d} \underbrace{\pi^{d/2} \cos(\lVert \bst \rVert)}_{=:g(\bst)} \cdot \underbrace{\frac{\exp(-\lVert \bst \rVert^2)}{\pi^{d/2}}}_{\text{density of } \calN(\bszero,\mathsf{I}/2)} \, \rmd \bst = \bbE(Y) = \bbE[g(\bsT)].
\end{equation}

Since the low discrepancy sequences that drive qMC are defined on the unit cube, $[0,1]^d$, we perform a variable transformation $\bst = \bigl (\Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d) \bigr )/\sqrt{2}$, where is $\Phi$ is the cumulative distribution function of the standard Gaussian random variable.  This reimagines the integral $I$ as the expectation of a function, $f$, of a standard uniform random variable:
\begin{multline}\label{eq:keisterB}
	\mu = \int_{[0,1]^d} \underbrace{\pi^{d/2} \cos\Bigl (\sqrt{2} \bigl \lVert \bigl( \Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d)\bigr) \bigr\rVert \Bigr)}_{=:f(\bsx)} \, \rmd \bsx \\
	= \bbE[Y]
	= \bbE[f(\bsX)], \qquad \bsX \sim \calU[0,1]^d.
\end{multline}

Consider the specific case of $d=6$ for which $\mu \approx -2.327303729298$.  We approximate $\mu$ by the sample mean defined in \eqref{eq:sample_mean} for three different kinds of nodes, $\bsX_0, \bsX_1, \ldots, \bsX_{n-1}$, for various sample sizes, $n$, and plot the relative errors, $\lvert (\mu - 	\hat{\mu}_n)/\mu\rvert$ in Figure \ref{fig:keister-err}. The three kind of nodes are
\begin{enumerate}
	\renewcommand{\labelenumi}{\roman{enumi}.}
	\item Cartesian grids, $\{1/(2m), \ldots, (2m-1)/(2m) \}^6$, for $m = 2, 3, \ldots$ (blue),
	\item IID simple random  with arbitrary $n$ (orange), and
	\item Randomized low discrepancy (Sobol') sequences with $n = 1, 2, \ldots, 2^m, \ldots $ (green).
\end{enumerate}

\begin{figure}
	\centering
	\includegraphics[width =10cm]{\figpath/n_is_16777216_d_is_6_n_rep_is_50_KeisterErrors.pdf}
	\caption{The relative error of approximationg $\mu$ defined in \eqref{eq:keisterB} by the sample mean, $\hat{\mu}_n$ defined in \eqref{eq:sample_mean} for various choices of nodes, $\bsX_0, \bsX_1, \ldots$.  Grids have the largest error and low discrepancy (qMC) nodes have the smallest error. \label{fig:keister-err}}
\end{figure}

Note the following from this example:
\begin{itemize}
	\item This integral is not particularly easy.  Even the best choice of nodes requires at least $n=100$ to get a relative error below $10\%$.

	\item Although grids may be attractive for low dimensional problems, e.g., $d = 1$, $2$, or $3$, they do poorly for this modest dimension, $d=6$.  For modest $d$, the possible sample sizes, $n$, are quite sparse, as is shown in Figure \ref{fig:keister-err}. Choosing the nodes to lie on a grid corresponds to a tensor product midpoint cubature rule, which would normally be expected to have an error of $\mathcal{O}(n^{-2/d})$  for general $d$.  The error decay of $\mathcal{O}(n^{-1/5})$ rather than $\mathcal{O}(n^{-2/6})$ for this example may be due to the lack of smoothness of $f$ near the boundaries of the unit cube.
	\item Simple MC (orange) are a substantial improvement over grid nodes.  Their root mean squared error is known to decay as $\mathcal{O}(n^{-1/2})$, as is observed here.  As a reminder, for IID nodes the root mean squared error
	\begin{equation}\label{eq:IIDerror}
		\sqrt{\mathbb{E}[(\mu - \hat{\mu})^2]} = \sqrt{\var(\hat{\mu})} = \sqrt{\frac{\var(f(\bsX))}{n}} = \frac{\mathrm{Std}(f(\bsX))}{n^{1/2}},
	\end{equation*}
	where $\var$ denotes the variance and $\mathrm{Std}$ the standard deviation.  For simple MC
\end{itemize}

\section{Low Discrepancy Sequence Constructions} \label{sec:construct}
\section{Discrepancy} \label{sec:discrepancy}
\section{Randomization} \label{sec:random}
\section{Stopping Criteria} \label{sec:stop}
\section{Reformulating Our Problem} \label{sec:reformulate}
\section{Ongoing Research} \label{sec:onging}
\section{Conclusion} \label{sec:conclusion}

\bibliographystyle{spmpsci}
\bibliography{FJH23,FJHown23}
